---
title: "maxPlot"
author: "AHMED CHOUKRI"
Version: "1.0"
---

## Brief introdution to maxPlot

Designed to generate, from normalized or not normalized data, dose-response curves, downstream steps involve: 
1. verifying density distribution; 
2. bimodal/normality/variance homogeneity assumptions test; 
3. ANOVA test to choose the appropriate model to draw the curve of dose response and  
4. generating the dose response curves with confidence intervals, the maximum response and EC50s graphs for each molecule, and the respective error bars; and 
5. for the case of independent measurements, the output can be assembled for statistical assessment, which in turn generates Ec50s and maximum values with the signification levels assigned by letters. 

## Verifying and installing the packages
```{r}
check_and_install_packages <- function(packages) {
  for(package in packages) {
    if(!require(package, character.only = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
}

packages <- c("readxl", "drda", "ggplot2", "dplyr", "gridExtra", "openxlsx", "car","FSA", "multcompView", "diptest")

check_and_install_packages(packages)

library(readxl)
library(drda)
library(ggplot2)
library(dplyr)
library(car)
library(diptest)
library(FSA)
library(multcompView)
library(openxlsx)
```

## 1. Raw Data file

Raw data file needs to have three columns ('Molecule'), ('Concentration'), ('Response').
For this version of maxPlot, data shouldn't have '+ control' or 'blank' modalities because it assumes that data to be adjusted.

```{r}
file_path <- "C:/Users/Choukri-A/Desktop/exp.xlsx"

data <- read_excel(file_path)
```

## 2. Data distribution overview (dip.test if for bimodality's test)
1. verifying density distribution; 
2. bimodal/normality/variance homogeneity assumptions test; 
```{r}

shapiro_result <- shapiro.test(data$Response)
print(shapiro_result)


dip_result <- dip.test(data$Response)
print(dip_result)


levene_result <- car::leveneTest(data$Response, group = data$Molecule)
print(levene_result)


shapiro_df <- data.frame(Test = "Shapiro-Wilk Test", Statistic = shapiro_result$statistic, p.value = shapiro_result$p.value)
dip_df <- data.frame(Test = "Hartigan's Dip Test", Statistic = dip_result$statistic, p.value = dip_result$p.value)
levene_df <- data.frame(Test = "Levene's Test", Statistic = levene_result[1, "F value"], p.value = levene_result[1, "Pr(>F)"])

distribution_overview <- rbind(shapiro_df, dip_df, levene_df)

wb_tests <- createWorkbook()

addWorksheet(wb_tests, "Distribution Overview")

writeData(wb_tests, "Distribution Overview", x = distribution_overview, startRow = 1, startCol = 1)

saveWorkbook(wb_tests, "Distribution_Tests.xlsx", overwrite = TRUE)


png("QQplot_Response.png", width = 800, height = 600)

qqnorm(data$Response, main = "Q-Q plot of Response values")
qqline(data$Response)

dev.off()

png("DensityPlot_Response.png", width = 800, height = 600)

plot(density(data$Response), main = "Density plot of Response values", xlab = "Response")

dev.off()
```

## 3. ANOVA test to choose the appropriate model to draw the curve of dose response and  
## 4. Generating the dose response curves with confidence intervals
This part is for extracting model parameters and their standard errors by creating a loop, watch out the excel file for model selection (4-parameters, 5 parameters or 2 parameters)
This is not ggplot based, it's the default plot() option by drda package, it plots the best model fit for a given molecule/germination stimulant with confidence interval highlighted.
'phi' which is the log10EC50, is used to define the threshold to draw the curve which is set at -1 (less than -1 means the curve can be drawn).
The latter step will create a dataframe withe values strictly less or equal -1 for the upcoming plot.
```{r}
molecules <- unique(data$Molecule)

summaries <- list()
anova_results <- list()
plots_list <- list()

max_responses <- rep(NA, length(molecules))
se_max_responses <- rep(NA, length(molecules))
ec50s <- rep(NA, length(molecules))
se_ec50s <- rep(NA, length(molecules))

for(i in seq_along(molecules)) {
  molecule <- molecules[i]
  
  subset_data <- subset(data, Molecule == molecule)

  subset_data <- subset_data %>%
    filter(Concentration > 0)
  
  fit <- drda(Response ~ log10(Concentration), data = subset_data, mean_function = "gz")

  params <- coef(fit)
  se_params <- sqrt(diag(vcov(fit)))
  
  ec50 <- 10^(params["phi"])  
  se_ec50 <- abs(se_params["phi"] * ec50 * log(10)) 

  max_response <- params["alpha"] + params["delta"]
  se_max_response <- sqrt(se_params["alpha"]^2 + se_params["delta"]^2)  # error propagation was used here

  if(params["phi"] <= -1) {  
    ec50s[i] <- ec50
    se_ec50s[i] <- se_ec50
  }
  max_responses[i] <- max_response
  se_max_responses[i] <- se_max_response
  
  file_name <- paste0("Dose_Responses_Plot_", molecule, ".png")
  
  png(file_name, width = 800, height = 600)

  if(params["phi"] > -1) {
    plot(log10(subset_data$Concentration), subset_data$Response, main = paste("Non-fitted case:", molecule), xlab = "log10(dose)",
         cex.main = 1.5, 
         cex.lab = 4,  
         cex.axis = 2, 
         col.axis = "blue", 
         col.lab = "red", 
         col.main = "darkgreen", 
         pch = 19, 
         col = "darkorange", 
         lwd = 5, 
         lty = 5) 
    
   
    grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted")
  } else {
    
    plot(fit, main = paste("Best model fit for", molecule), xlab = "log10(dose)",
         cex.main = 1.5, 
         cex.lab = 4,  
         cex.axis = 2, 
         col.axis = "blue", 
         col.lab = "red", 
         col.main = "darkgreen", 
         pch = 19, 
         col = "darkorange", 
         lwd = 5, 
         lty = 5) 
    
    
    grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted")
  }
  
  dev.off()
  
  plots_list[[molecule]] <- file_name 
}

```

# 5. Saving the numerical results in excel file on the working directory
Do not change the name of the files generated

```{r}
results <- data.frame(
  Molecule = molecules, 
  MaxResponse = max_responses, 
  SE_MaxResponse = se_max_responses, 
  EC50 = ec50s, 
  SE_EC50 = se_ec50s
)

print(results)

wb <- createWorkbook()

addWorksheet(wb, "Results")

writeData(wb, "Results", results, startRow = 1, startCol = 1)

for(i in seq_along(molecules)) {
  addWorksheet(wb, paste("Summary_", molecules[i]))
  writeData(wb, paste("Summary_", molecules[i]), summaries[[molecules[i]]])    
  addWorksheet(wb, paste("ANOVA_", molecules[i]))
  writeData(wb, paste("ANOVA_", molecules[i]), anova_results[[molecules[i]]])
  writeData(wb, paste("Summary_", molecules[i]), startCol = 10, startRow = 1, plots_list[[molecules[i]]])
}

filename <- tools::file_path_sans_ext(basename(file_path))

output_file <- paste0("Dose_Response_Summary_ANOVA_Plots_", filename, ".xlsx")

saveWorkbook(wb, output_file, overwrite = TRUE)
```

# 6. Generating the maximum response and EC50s graphs for each molecule, and the respective error bars
The generated graphs aren't that cool but they show clearly the variations amongest the molecules.
```{r}
ec50_plot <- ggplot(results[!is.na(results$EC50), ], aes(x = EC50, y = Molecule)) + 
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_errorbar(aes(xmin = EC50 - SE_EC50, xmax = EC50 + SE_EC50), width = 0.2) +
  scale_x_continuous(trans = 'log10', breaks = 10^seq(-14, -1, by = 1)) + 
  coord_cartesian(xlim = c(10^-14, 10^-1)) +  
  labs(title = "EC50 Values", x = "EC50 (log scale)", y = NULL) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(hjust = 1),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(size = 12)) +  
  scale_y_discrete(position = "right")

ggsave("EC50_Plot.png", ec50_plot, width = 10, height = 8, units = "in")

max_response_plot <- ggplot(results, aes(x = Molecule, y = MaxResponse)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  geom_errorbar(aes(ymin = MaxResponse - SE_MaxResponse, ymax = MaxResponse + SE_MaxResponse), width = 0.2) +
  labs(title = "Maximum Responses", x = NULL, y = "Maximum Response") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.justification = "top",
        legend.box.just = "right")

ggsave("Max_Response_Plot.png", max_response_plot, width = 10, height = 8, units = "in")

grid.arrange(ec50_plot, max_response_plot, nrow = 2)
```

# 7. Assemble independant measurements for statitsics
For the case of independent measurements, the output can be assembled for statistical assessment, which in turn generates Ec50s and maximum values with the signification levels assigned by letters.
```{r eval=FALSE, echo=FALSE}

files <- list.files(pattern = "^Dose_Response_Summary_ANOVA_Plots")

all_results <- data.frame()

for(file in files) {
  results <- read_excel(file, sheet = "Results")
  
  results$Molecule <- as.factor(results$Molecule)
  
  all_results <- rbind(all_results, results)
}

anova_result_EC50 <- aov(EC50 ~ Molecule, data = all_results)
anova_result_MaxResponse <- aov(MaxResponse ~ Molecule, data = all_results)

tukey_result_EC50 <- TukeyHSD(anova_result_EC50)
tukey_result_MaxResponse <- TukeyHSD(anova_result_MaxResponse)

group_letters_EC50 <- multcompLetters(tukey_result_EC50$Molecule[, "p adj"])
group_letters_MaxResponse <- multcompLetters(tukey_result_MaxResponse$Molecule[, "p adj"])

results_df <- data.frame(
  Molecule = levels(all_results$Molecule),
  EC50_p_value = tukey_result_EC50$Molecule[, "p adj"],
  MaxResponse_p_value = tukey_result_MaxResponse$Molecule[, "p adj"],
  EC50_rank = group_letters_EC50$Letters,
  MaxResponse_rank = group_letters_MaxResponse$Letters
)

all_results <- merge(all_results, results_df, by = "Molecule")

aggregated_results <- all_results %>%
  group_by(Molecule) %>%
  summarise(
    Count = n(),
    Mean_EC50 = mean(EC50, na.rm = TRUE),
    SD_EC50 = ifelse(n() > 1, sd(EC50, na.rm = TRUE), 0),
    SE_EC50 = ifelse(n() > 1, sd(EC50, na.rm = TRUE) / sqrt(n()), 0),
    Mean_MaxResponse = mean(MaxResponse, na.rm = TRUE),
    SD_MaxResponse = ifelse(n() > 1, sd(MaxResponse, na.rm = TRUE), 0),
    SE_MaxResponse = ifelse(n() > 1, sd(MaxResponse, na.rm = TRUE) / sqrt(n()), 0),
    EC50_rank = first(EC50_rank),
    MaxResponse_rank = first(MaxResponse_rank)
  )

print(aggregated_results)

ec50_plot <- ggplot(aggregated_results, aes(x = Mean_EC50, y = Molecule)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_errorbar(aes(xmin = Mean_EC50 - SE_EC50, xmax = Mean_EC50 + SE_EC50), width = 0.2) +
  geom_text(aes(label = EC50_rank), hjust = -0.5, size = 4) +  
  scale_x_continuous(trans = 'log10', breaks = 10^seq(-14, -1, by = 1)) +  
  coord_cartesian(xlim = c(10^-14, 10^-1)) +  
  labs(title = "EC50 Values -Tukey's HSD test", x = "EC50 (log scale)", y = NULL) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(hjust = 1),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(size = 12)) +  
  scale_y_discrete(position = "right")

ggsave("EC50_Plot1.png", ec50_plot, width = 10, height = 8, units = "in")

max_response_plot <- ggplot(aggregated_results, aes(x = Molecule, y = Mean_MaxResponse)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  geom_errorbar(aes(ymin = Mean_MaxResponse - SE_MaxResponse, ymax = Mean_MaxResponse + SE_MaxResponse), width = 0.2) +
  geom_text(aes(label = MaxResponse_rank), vjust = -0.3, hjust = -0.5, size = 4) +  
  labs(title = "Maximum Responses", x = NULL, y = "Maximum Response") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.justification = "top",
        legend.box.just = "right")

ggsave("Max_Response_Plot1.png", max_response_plot, width = 10, height = 8, units = "in")


grid.arrange(ec50_plot, max_response_plot, nrow = 2)

tukey_df_EC50 <- data.frame(tukey_result_EC50$Molecule)
tukey_df_MaxResponse <- data.frame(tukey_result_MaxResponse$Molecule)

wb <- createWorkbook()

addWorksheet(wb, "Results")
writeData(wb, "Results", results_df)

addWorksheet(wb, "Tukey Summary EC50")
writeData(wb, "Tukey Summary EC50", tukey_df_EC50)

addWorksheet(wb, "Tukey Summary MaxResponse")
writeData(wb, "Tukey Summary MaxResponse", tukey_df_MaxResponse)
addWorksheet(wb, "Aggregated Results")
writeData(wb, "Aggregated Results", aggregated_results)

saveWorkbook(wb, "Final_statistics.xlsx", overwrite = TRUE)

```
